Innovations Report: Procedural Fur System

### 2018-03-18 docs/paper.tex
more on docs

\section{Introduction} \label{sec:introduction}
Introduction Text.
\subsection{Related Work} \label{sec:related}
nvidia hairworks

\subsection{Existing Solutions} \label{sec:existing}
Whilst many larger companies opt to develop their own software for handling fur, such as MPC's Furtility \cite{furtility}, there are a number of commercially available solutions that are also widely used in production.

The most popular of these is XGen, originally developed by Walt Disney Animation Studios \cite{xgen}, which has been included in Autodesk's Maya since 2014. XGen uses a 'Collection' and 'Description' paradigm; Descriptions store user controlled parameters which affect fur primitives (curves for example) on a select area of a mesh, and Collections act as a container for these Descriptions. There are two primary methods offered for distributing primitives on a mesh, Guide Curves can be used to manually control the placement and shape of primitives by placing 'Guides' on the mesh surface, whereas 'Groomable Splines' randomly distirbutes splines over the mesh surface which can be manipulated in real time using viewport manipulators that mimic the brushing and styling of fur. In both modes, 'modifiers' can be added to the modifier stack to further adjust the visual outcome, adding features such as clumping, noise, and coiling. Attributes within these modifiers can be further controlled using expressions or texture maps.

Another commercially available solution is Peregrine*Labs' Yeti \cite{yeti}, which is also offered as a plugin within Autodesk's Maya. Yeti offers a more procedural approach, giving the user a node graph interface in which to construct the desired look from individual components. For example, the most simple usage requires the user to first create a 'scatter' node to distribute points on the mesh surface, and then combine it with a 'grow' node to extrude curves from these points. Whilst at first this may seem unintuitive when compared to XGen, it offers greater flexibility and avoids the pitfalls associated with a destructive workflow.

@@

\subsection{GPU Acceleration} \label{sec:existingGPU}
Whilst it is not clear to what extent each of these solutions make use of GPU acceleration, they all perform fast enough to allow the user to manipulate them interactively. I therefore decided to monitor the CPU and GPU usage using \verb|htop| and \verb|nvidia-smi| respectively, before and after adding XGen content to a simple maya scene, to determine where the majority of data was being computed.

@@

I wanted to include a node-graph style interface within my application, as it is commonly used within existing artist tools (Autodesk Maya, Unreal Engine), and would encourage modularity within my API. Qt does not natively provide this kind of interface, so I made use of NodeEditor \cite{Pinaev2017}, an existing Qt-based library that provides this functionality.

@@

\subsubsection{Mesh Loading} \label{sec:objLoading}
% The first problem I tackled was that of loading mesh geometry into the system, and storing it in a Shader Storage Buffer Object (SSBO) for later use in a compute shader. I opted to implement a Wavefront OBJ loader as they are (deceptively) simple to parse, consisting of a list of vertex positions, vertex normals, vertex texture coordinates, and faces defined as a list of indices. I created an OBJ loader node using the NodeEditor library, which parses

@@

\subsubsection{Curve Distributors} \label{sec:distributors}

@@

There are many different methods one could use to distribute points on an abritrary mesh, however for replicating creature fur it is desirable to choose a method that distributes primitives at random positions across the surface, yet roughly evenly spaced from neighboring primitives. A popular method used to achieve this kind of distribution is referred to as 'Dart Throwing', which relies on checking the distance to neighboring samples after a new sample has been generated, and discarding the sample if the samples are too close. Whilst these kinds of methods produce good results, they can be inefficient as in certain situations many samples may be generated and rejected, before a sample is accepted. An optimisation for this is described in \textit{Dart Throwing on Surfaces} \cite{dartThrowing}, which relies on removing areas of the mesh that are already covered by a sample, by means of triangle subdivision.

Unfortunately, parallelizing dart thowing methods is not trivial, making them unsuitable for my desired usage of OpenGL compute shaders. Whilst sampling methods suitable for parallel processing do exist, such as those described in \textit{Parallel Poisson Disk Sampling with Spectrum Analysis on Surfaces} \cite{parallelPoisson}, I instead opted to implement a naive random samling method as it was much easier to parallelize.

@@

\subsubsection{Compute Shaders} \label{sec:compute shaders}

In order to get this distribution process running on the GPU, I needed to allocate and store data for both the Mesh and Curves in OpenGL buffers, for later use in a compute shader. The most common way of storing OBJ data in OpenGL buffers, as described in \cite{objLoading}, is to use a \verb|GL_ARRAY_BUFFER| object to store the vertex data and a \verb|GL_ELEMENT_ARRAY_BUFFER| to store the indices of the \verb|GL_ARRAY_BUFFER| that make up each face. However for ease of use in compute shaders I decided instead to store each face in a \verb|GL_SHADER_STORAGE_BUFFER|, where each face consists of three sets of positions, normals, and texture coordinates. Whilst this is less efficient, I found that the overhead in GPU memory usage was negligible, and that this method allowed for much simpler usage in a compute shader. Allocating a buffer for the Curves data was much simpler as it consisted of an array of 'Curve' structs, each consisting of 5 vertices. Listing~\ref{lst:buffers} shows how these buffers were accessed from within the compute shaders; note how all of the \texttt{vec3} values have been padded to \texttt{vec4}, as the \texttt{std430} layout qualifier requires an alignment of 16 bytes.

\subsubsection{Curve Operators} \label{sec:operators}

@@

Whilst rendering and shading triangulated meshes is relatively straightforward, as described in \cite{objLoading}, there are multiple different approaches one could use to render curves. The technique I chose to implement relies on drawing an empty VAO multiple times, and using \texttt{gl\_PrimitiveIDIn} from within a geometry shader to determine which curve from the bound \texttt{GL\_SHADER\_STORAGE\_BUFFER} should be drawn. Listings~\ref{lst:rendercpp} and ~\ref{lst:renderglsl} shows how this was achieved.

@@

\subsection{User Testing} \label{sec:usertesting}
Near the end of the development I was able to convince friend and fellow student Alin Bolcas to test the system and provide feedback from an artist's point of view; Alin is an experienced modelling and look development artist who uses a variety of different software packages to create his work, so getting his feedback on the usability of my tool proved extremely useful.

@@

\begin{figure}[htbp]\centering
\includegraphics[width=1.0\linewidth]{images/test_0007}
\caption{\label{figure:screenshot} An example of the visual output that the system is capable of producing.}
\end{figure}

@@

\section{Future Work} \label{sec:futurework}

>>>>

\section{Introduction} \label{sec:introduction}
Introduction Text.
\subsection{Related work} \label{sec:related}
nvidia hairworks

\subsection{Existing solutions} \label{sec:existing}
Whilst many larger companies opt to develop their own software for handling fur, such as MPC's Furtility~\cite{furtility}, there are a number of commercially available solutions that are also widely used in production.

The most popular of these is XGen, originally developed by Walt Disney Animation Studios~\cite{xgen}, which has been included in Autodesk's Maya since 2014. XGen uses a 'Collection' and 'Description' paradigm; Descriptions store user controlled parameters which affect fur primitives (curves for example) on a select area of a mesh, and Collections act as a container for these Descriptions. There are two primary methods offered for distributing primitives on a mesh, Guide Curves can be used to manually control the placement and shape of primitives by placing 'Guides' on the mesh surface, whereas 'Groomable Splines' randomly distirbutes splines over the mesh surface which can be manipulated in real time using viewport manipulators that mimic the brushing and styling of fur. In both modes, 'modifiers' can be added to the modifier stack to further adjust the visual outcome, adding features such as clumping, noise, and coiling. Attributes within these modifiers can be further controlled using expressions or texture maps.

Another commercially available solution is Peregrine*Labs' Yeti~\cite{yeti}, which is also offered as a plugin within Autodesk's Maya. Yeti offers a more procedural approach, giving the user a node graph interface in which to construct the desired look from individual components. For example, the most simple usage requires the user to first create a 'scatter' node to distribute points on the mesh surface, and then combine it with a 'grow' node to extrude curves from these points. Whilst at first this may seem unintuitive when compared to XGen, it offers greater flexibility and avoids the pitfalls associated with a destructive workflow.

@@

\subsection{GPU acceleration} \label{sec:existingGPU}
Whilst it is not clear to what extent each of these solutions make use of GPU acceleration, they all perform fast enough to allow the user to manipulate them interactively. I therefore decided to monitor the CPU and GPU usage using \verb|htop| and \verb|nvidia-smi| respectively, before and after adding XGen content to a simple maya scene, to determine where the majority of data was being computed.

@@

I wanted to include a node-graph style interface within my application, as it is commonly used within existing artist tools (Autodesk Maya, Unreal Engine), and would encourage modularity within my API. Qt does not natively provide this kind of interface, so I made use of NodeEditor~\cite{Pinaev2017}, an existing Qt-based library that provides this functionality.

@@

\subsubsection{Mesh loading} \label{sec:objLoading}
% The first problem I tackled was that of loading mesh geometry into the system, and storing it in a Shader Storage Buffer Object (SSBO) for later use in a compute shader. I opted to implement a Wavefront OBJ loader as they are (deceptively) simple to parse, consisting of a list of vertex positions, vertex normals, vertex texture coordinates, and faces defined as a list of indices. I created an OBJ loader node using the NodeEditor library, which parses

@@

\subsubsection{Curve distributors} \label{sec:distributors}

@@

There are many different methods one could use to distribute points on an abritrary mesh, however for replicating creature fur it is desirable to choose a method that distributes primitives at random positions across the surface, yet roughly evenly spaced from neighboring primitives. A popular method used to achieve this kind of distribution is referred to as 'Dart Throwing', which relies on checking the distance to neighboring samples after a new sample has been generated, and discarding the sample if the samples are too close. Whilst these kinds of methods produce good results, they can be inefficient as in certain situations many samples may be generated and rejected, before a sample is accepted. An optimisation for this is described in \textit{Dart Throwing on Surfaces}~\cite{dartThrowing}, which relies on removing areas of the mesh that are already covered by a sample, by means of triangle subdivision.

Unfortunately, parallelizing dart thowing methods is not trivial, making them unsuitable for my desired usage of OpenGL compute shaders. Whilst sampling methods suitable for parallel processing do exist, such as those described in \textit{Parallel Poisson Disk Sampling with Spectrum Analysis on Surfaces}~\cite{parallelPoisson}, I instead opted to implement a naive random sampling method as it was much easier to parallelize.

@@

\subsubsection{Compute shaders} \label{sec:compute shaders}

In order to get this distribution process running on the GPU, I needed to allocate and store data for both the Mesh and Curves in OpenGL buffers, for later use in a compute shader. The most common way of storing OBJ data in OpenGL buffers, as described in~\cite{objLoading}, is to use a \verb|GL_ARRAY_BUFFER| object to store the vertex data and a \verb|GL_ELEMENT_ARRAY_BUFFER| to store the indices of the \verb|GL_ARRAY_BUFFER| that make up each face. However for ease of use in compute shaders I decided instead to store each face in a \verb|GL_SHADER_STORAGE_BUFFER|, where each face consists of three sets of positions, normals, and texture coordinates. Whilst this is less efficient, I found that the overhead in GPU memory usage was negligible, and that this method allowed for much simpler usage in a compute shader. Allocating a buffer for the Curves data was much simpler as it consisted of an array of 'Curve' structs, each consisting of 5 vertices. Listing~\ref{lst:buffers} shows how these buffers were accessed from within the compute shaders; note how all of the \texttt{vec3} values have been padded to \texttt{vec4}, as the \texttt{std430} layout qualifier requires an alignment of 16 bytes.

\subsubsection{Curve operators} \label{sec:operators}

@@

Whilst rendering and shading triangulated meshes is relatively straightforward, as described in~\cite{objLoading}, there are multiple different approaches one could use to render curves. The technique I chose to implement relies on drawing an empty VAO multiple times, and using \texttt{gl\_PrimitiveIDIn} from within a geometry shader to determine which curve from the bound \texttt{GL\_SHADER\_STORAGE\_BUFFER} should be drawn. Listings~\ref{lst:rendercpp} and ~\ref{lst:renderglsl} shows how this was achieved. The use geometry shaders also allowed me to alternate between drawing the curves using line primitives, or to draw the curves as forward facing triangle strips. % which I refer to as 'ribbons'.

For shading of the meshes I opted to implement a physically based shading model, as decribed in~\cite{learnPBR}, however some alterations were necessary as my method of rendering curves does not provide usable surface normal information. Instead, the normalized direction in which the curve is leading from one vertex to the next, was substituted as the surface normal. Then, all of the dot product terms in the shading calculations which use the surface normal, were inversed. The result roughly approximates the shading qualities of thin cylinders by assuming that all vectors that are perpendicular to the direction of the curve are valid surface normals. Whilst this technique provides visually pleasing results, it is inaccurate and does not account for properties such as self-shadowing, or light scattering.

@@

\subsection{User testing} \label{sec:usertesting}
Near the end of the development I was able to convince friend and fellow student Alin Bolcas to test the system and provide feedback from an artist's point of view; Alin is an experienced modelling and look development artist who uses a variety of different software packages to create his work, so getting his feedback on the usability of my tool proved extremely useful.

@@

After I had implemented some of the features Alin requested, I decided that the system was in a satisfactory state for the purposes of this assignment. The following figure~\ref{figure:screenshots} shows some memes

\begin{figure}[htbp]
\centering
\subfloat[]{\includegraphics[width=0.45\linewidth]{images/rick}}
\hfill
\subfloat[]{\includegraphics[width=0.45\linewidth]{images/bunny}}
\caption{\label{figure:screenshots} literal memes}
\end{figure}

\begin{figure}[htbp]\centering
\includegraphics[width=1.0\linewidth]{images/test_0007}
\caption{\label{figure:screenshot1} An example of the visual output that the system is capable of producing.}
\end{figure}

@@

\subsection{Future Work} \label{sec:futurework}
